A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-04-18 paper.tex

Figures \ref{fig:polypower} and \ref{fig:heterpower} present the power
curves of various tests plotted against the effect size in the
residuals, for non-linearity and heteroskedasticity, respectively. In
each case the power of visual test is re-calculated for multiple
bootstrap samples leading to the many (orange) curves. The effect size
was computed at a 5\% significance level and plotted on a natural
logarithmic scale. To facilitate visual calibration of effect size
values with the corresponding diagnostic plots, a sequence of example
residual plots with increasing effect sizes is provided at the bottom of
these figures. These plots serve as a visual aid to help readers
understand how different effect size values translate to changes in the
diagnostic plots. The horizontal lines of dots at 0 and 1 represent the
non-reject or reject made by visual tests for each lineup.

@@

Figures \ref{fig:polypower} compares the power for the different tests
for non-linear structure in the residuals. The test with the uniformly
higher power is the RESET test, one that specifically tests for
non-linearity. The power curves for the visual test are effectively a
shift right from that of the RESET test. This means that the RESET test
will reject a lower effect size (less structure) than the visual test,
but otherwise the performance will be similar. In other words, the RESET
test is more sensitive than the visual test. This is not necessarily a
good feature, for the purposes of diagnosing model defects. If one scans
the residual plot examples at the bottom, we might argue that the
non-linearity is not sufficiently problematic until an effect size of
around 3 or 3.5. The RESET test would reject closer to an effect size of
2, but the visual test would reject closer 3.25. As expected the BP and
SW tests have much lower power - they are not designed to detect
non-linearity.

For the heteroskedasticity pattern, the power of BP test, designed for
detecting heteroskedasticity, is uniformly higher than the other tests.
The visual test power curve is a right shift. This shows a similar story
to the power curves for non-linearity pattern - the conventional test is
more sensitive than the visual test. From the example residual plots at
the bottom we might argue that the heteroskedasticity becomes noticeably
visible around an effect size of 3 or 3.5. However the BP test would
reject at around effect size 2.5. Interestingly, the power curve for the
SW test (for non-normality) is only sighly different to that of the
visual test, suggesting that it performs reasonably for detecting
heteroskedasticity, too. The power curve for the BP test suggests it is
not useful for detecting heteroskeadsticity, as expected.

@@

@@

Now it is important to note that residual plots need to be delivered as
a lineup, where it is embedded in a field of null plots. A residual plot
may contains many visual features, but some are caused by the
characteristics of the predictors and the randomness of the error, not
by the violation of the model assumptions. These irrelevant visual
features have a chance to be filtered out by subjects with a comparison
to null plots, results in a set of more accurate visual findings. This
enables a careful calibration for reading structure in residual plots.

However, human evaluation of residuals is expensive. It is
time-consuming, laborious and unfriendly to vision-impaired people. This
is another reason why it often appears to be ignored. With the
availability of sophisticated computer vision algorithms today, the goal
of this work is to form the basis of providing automated residual plot
reading. The findings suggest the strong demand of graphical inspection
in regression diagnostics, so developing an automatic visual inference
system to evaluate lineups of residual plots is valuable. We plan to
build a completed open-source system in an \texttt{R} package and
provide a web interface such as a website for public to interact with.
Details about this system will be discussed in our next paper.

The experiment also revealed some interesting details about how residual
plots are read. For the most part, the visual test performed very
similarly to the appropriate conventional test only with the power curve
shifted in the less sensitive direction. Unlike the conventional tests,
where one needs to specifically test for non-linearity or
heteroskedasticity the visual test operated effectively across the range
of departures from good residuals.

As expected, if the fitted value distribution is not uniform, there is a
loss of power in the visual test. Structure is hardest to detect if
fitted values are lognormal. Also, complex structure are generally
harder to detect, but there are outliers. Under the designed scenarios
in this paper, we find the visual test to be a more robust test against
the change of fitted value distributions. A surprising finding was that
the direction of heteroskedasticity appears to affect the ability to
visually detect it, with wedge to the right being less detectable.

>>>>

Figures \ref{fig:polypower} and \ref{fig:heterpower} present the power
curves of various tests plotted against the effect size in the
residuals, for non-linearity and heteroskedasticity, respectively. In
each case the power of visual test is calculated for multiple bootstrap
samples leading to the many (orange) curves. The effect size was
computed at a 5\% significance level and plotted on a natural
logarithmic scale. To facilitate visual calibration of effect size
values with the corresponding diagnostic plots, a sequence of example
residual plots with increasing effect sizes is provided at the bottom of
these figures. These plots serve as a visual aid to help readers
understand how different effect size values translate to changes in the
diagnostic plots. The horizontal lines of dots at 0 and 1 represent the
non-reject or reject made by visual tests for each lineup.

@@

Figure \ref{fig:polypower} compares the power for the different tests
for non-linear structure in the residuals. The test with the uniformly
higher power is the RESET test, one that specifically tests for
non-linearity. (Note that the BP and SW tests have much lower power,
which is expected because they are not designed to detect
non-linearity.) The bootstrapped power curves for the visual test are
effectively a shift right from that of the RESET test. This means that
the RESET test will reject at a lower effect size (less structure) than
the visual test, but otherwise the performance will be similar. In other
words, the RESET test is more sensitive than the visual test. This is
not necessarily a good feature for the purposes of diagnosing model
defects. If one scans the residual plot examples at the bottom, we might
argue that the non-linearity is not sufficiently problematic until an
effect size of around 3 or 3.5. The RESET test would reject closer to an
effect size of 2, but the visual test would reject closer 3.25, for a
significance level of 0.05.

For the heteroskedasticity pattern, the power of BP test, designed for
detecting heteroskedasticity, is uniformly higher than the other tests.
The visual test power curve is a right shift. This shows a similar story
to the power curves for non-linearity pattern - the conventional test is
more sensitive than the visual test. From the example residual plots at
the bottom we might argue that the heteroskedasticity becomes noticeably
visible around an effect size of 3 or 3.5. However the BP test would
reject at around effect size 2.5. Interestingly, the power curve for the
SW test (for non-normality) is only slightly different to that of the
visual test, suggesting that it performs reasonably for detecting
heteroskedasticity, too. The power curve for the BP test suggests it is
not useful for detecting heteroskeadsticity, as expected.

@@

Because the power curve of the visual tests are a shift to the right of
the conventional test (Figures \ref{fig:polypower} and
\ref{fig:heter-example}) it would be curious to know whether adjusting
the significance level (to .001, .0001, .00001, \ldots) of the
conventional test would generate similar decisions to that of the visual
test. Interestingly, it doesn't. Despite resulting less rejections,
neither the RESET or BP tests come to complete agreement with the visual
test (See Appendix).

@@

Now it is important to note that residual plots need to be delivered as
a lineup, where it is embedded in a field of null plots. A residual plot
may contain many visual features, but some are caused by the
characteristics of the predictors and the randomness of the error, not
by the violation of the model assumptions. These irrelevant visual
features have a chance to be filtered out by subjects with a comparison
to null plots, resulting in more accurate reading. The lineup enables a
careful calibration for reading structure in residual plots.

Human evaluation of residuals is expensive, though. It is time-consuming
and laborious. This is possibly why residual plot analysis is often not
done, in practice. However, with the emergence of effective computer
vision this work could form the basis to provide automated residual plot
reading.

The experiment also revealed some interesting results. For the most
part, the visual test performed very similarly to the appropriate
conventional test except that it has a power curve shifted in the less
sensitive direction. Unlike the conventional tests, where one needs to
specifically test for non-linearity or heteroskedasticity the visual
test operated effectively across the range of departures from good
residuals. If the fitted value distribution is not uniform, there is a
small loss of power in the visual test. But quite a surprise to us was
the big difference in power of the conventional test, which appear to be
quite sensitive to how the predictors are sampled. Another surprising
finding was that the direction of heteroskedasticity appears to affect
the ability to visually detect it: both triangles being more difficult
to detect than the butterfly, and a small difference in detection
between left- and right-triangle.
